Procurements of the Canadian Government, Runaway Military Contracts

### 2024-12-03 paper/paper.qmd
editing

---
title: "Procurements of the Canadian Government, Runaway Military Contracts"
author: 
  - Robert Ford
thanks: "Code and data are available at: https://github.com/Ford-Robert"
date: today
date-format: long
abstract: "TODO First sentence. TODO Second sentence. TODO Third sentence. TODO Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

In recent years, Canada's federal spending landscape has undergone significant transformations, particularly with a marked increase in military expenditures. The government has announced substantial new military spending packages, [REF]

Federal procurements involve the acquisition of goods and services by government departments, Buyers, from companies, Suppliers. These procurements are essential for a wide range of government functions from; acquiring defense equipment, building infrastructure, procuring healthcare supplies, technological services and engineering consultants for example. This paper will analyse federal procurement data to dissect government procurements over the last four years.

@@

I collected and processed detailed data on over 301,000 contracts awarded since January 2020. The dataset includes variables such as contract amounts, durations, award dates, Buyers, and Suppliers. We derived additional metrics to facilitate our analysis, such as Baseline Days (the number of days since January 1, 2020) and Duration Days (the length of each contract), and standardized supplier names to ensure consistency.

Some significant findings presented in this paper include: military procurement spending has surged in the past few years. Where the Department of National Defence now dominates federal procurement spending, with expenditures exceeding \$56 billion since 2020. This increase in spending is in part thanks to large contracts, including a \$11.2 billion contract to SkyAlyne for fighter pilot training—the largest in our dataset—and multi-billion-dollar contracts for fighter jets and naval vessels.

While  other government departments have experienced a notable decrease in procurement spending. For instance, the Public Health Agency of Canada's spending, which was substantial in 2020 due to the COVID-19 pandemic, has declined significantly in subsequent years.

@@

The paper is structured as follows: [Section 2] details the data collection and processing methodologies, including the sources of procurement data and the steps taken to clean and standardize the dataset. Then I present a series of graphs to demonstrate government procurements from various angles. [Section 3] describes the linear regression models used to assess trends in contract amounts over time and contract duration. [Section 4] presents the results of the analysis, highlighting key trends in military and non-military procurement spending. [Section 5] discusses the limitations of the study, including data constraints and methodological considerations, and suggests areas for future research. Finally, Section 6 concludes with a summary of the findings and their implications for policymakers, industry stakeholders, and future studies.

By providing a comprehensive analysis of federal procurement spending, this paper aims to fill the existing gap in understanding how government consumption of private sector goods and services is changing over time. The insights gained from this study can inform policy decisions, contribute to academic discourse, and enhance transparency regarding government spending practices.

TODO Cite R and all other packages

TODO Estimand?

@@

The data was collected using R and other packages [REF]. A web scraping script was used to collect all Federal Award Data from the IJF [REF]. The variables this paper focuses collected from the IJF are; Contract, Buyer, Supplier, Amount, Award Date, Start Date, and End Date.

@@

The head of the IFJ Database:
```{r fig-head1}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Sample of the Data
#| fig-align: center
#| fig-pos: "H"

@@

Additional Derived Variables:

@@

The data was then cleaned. First I began by setting each variable to the correct data type. Then I removed any data point where Amount was negative or zero, because I am not sure what a negative amount means in this context and contracts that are not worth anything are not worth looking at. There were a number of contracts which started and ended on the same day, to simplify later calculation I simply added a day to their End Date, figuring that same day contracts are equivalent to one day contracts. Further data cleaning details are available in Appendix TODO. After cleaning I was left with 301,094 contracts spanning TODO to TODO.

@@

The data is recorded and published through various government agencies, who use different accounting standards and procedures. The three sources IFJ used to collect the federal procurement data are; Canada Buys, Buy and Sell, and Proactive Disclosures (which makes up the vast majority of the procurement data). However, because the data used in this paper is pulled from the IFJ’s Federal Awards page this data comes exclusively from the Proactive Disclosures. The Treasury Board of Canada Secretariat is responsible for publishing these Proactive Disclosures under the Access to Information Act [REF].

@@

This is a break down of total amount awarded by the top 5 biggest spending Buyers. The Other category consists the of the total awards given by the remaining 88 smallest Government Buyers.  It is clear that National Defence dwarfs the procurement budget, spending over \$56 Billion on procurements since 2020.

@@

FIG show how National Defence spending has taken off in recent years, where is 2021 it had not even crested \$5 Billion, but in the last 3 years spending has exploded to \$15 Billion a year.

@@

# Create the bar chart
ggplot(data = military_spending_per_year, aes(x = year, y = total_spending)) +
  geom_bar(stat = "identity", fill = def_c) +
  labs(x = "Year", y = "Total Spending (Billions)") +
  theme_minimal() +
  scale_y_continuous() +
  theme(plot.title = element_text(hjust = 0.5))

@@

Public Services and Procurements Canada is the clearinghouse for the entire federal government. This means they are responsible for managing the procurement process for buyers across the government [REF]. It is important to note that the PSPC sometimes does not publish small contracts, less than \$40,000.  PSPC spending over time shows while military spending increases the rest of the government is taking a hit.
```{r fig-public-spending}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Public Services Spending Per Year in Billions
#| fig-align: center
#| fig-pos: "H"

@@

# Create the bar chart
ggplot(data = public_spending_per_year, aes(x = year, y = total_spending)) +
  geom_bar(stat = "identity", fill = public_c) +
  labs(x = "Year", y = "Total Spending (Billions)") +
  theme_minimal() +
  scale_y_continuous() +
  theme(plot.title = element_text(hjust = 0.5))

@@

This FIG shows the top five Buyers spending per year since 2020. In the first 2 years military spending held steady and did not dominate the budget by any means. But since then Defence spending exploded not just nominally but compared to the rest of the budget as well. Notably the Public Health Agency of Canada’s spending while strong in 2020, decreased to nothing over the following four years. The spending in 2020 is linked to purchases of medical equipment, so it is possible they have not needed to invest in new equipment since then. But this illustrates a boarder trend among all departments except National Defence, where their spending has steadily declined in the last 4 years.

@@

To find out the firms that are benefiting the most through this procurement process we can look to FIG. Unsurprisingly, the biggest suppliers are primarily are awarded contracts by the military. Interestingly the top supplier, Skyalyne, was only awarded one huge 11.2 Billion contract for fighter pilot training services. This is the largest contract in the dataset. Then considering the largest contracts awarded in the last 3 years to; the F-35 program (Fighter Planes, 4.5 Billion), General Atomics (Airplanes, 1.7 Billion), CAE (Aviation Engineering and Training, 4.4 Billion), Airbus (Airplanes, 3.7 Billion), for a grand total of 25.5 Billion. These large single contracts devoted to advanced fighter jets and training pilots to fly them is part of the reason that military spending has exploded in the last 3 years. The military also paid Vancouver Shipyards 3.4 Billion in 2024 for Ships and Boat parts.

@@

Dettwiler is an engineering firm that provides the Canadian Space Agency with engineering and scientific consulting. Surprisingly 86.3% of the Space Agency’s procurement budget goes to Dettwiler.

## Distribution

@@

Though most contracts are small contracts (less than \$100,000), the sum of these 246,000 contracts is only worth a measly 7.34 Billion. The Skyalyne contract itself outweighs all the small contracts.

@@

# Model

@@

Tidyverse: For data manipulation and visualization [REF]

Dplyr: [REF]

Car: [REF]

Caret:[REF]

@@

TODO

@@

Amount:

Baseline Days:

Duration Days:

@@

# Results

@@

# Discussion

## First discussion point {#sec-first-point}

@@

Data cleaning was a critical step. Contracts with negative or zero amounts were excluded due to their ambiguous nature. Contracts that started and ended on the same day were adjusted by extending the end date by one day to simplify duration calculations. After these refinements, the dataset comprised 301,094 contracts spanning from [TODO: start date] to [TODO: end date].

@@

## Second discussion point

@@

## Third discussion point

@@

# Appendix {-}

# Web Scraping

@@

# Additional data details

@@

# Proccessing Supplier

TODO processed_supplier
Talk about what I did thourougly, see data_cleaning script

@@

# Model details {#sec-model-details}

>>>>

---
title: "Procurements of the Canadian Government, Runaway Military Contracts"
author: 
  - Robert Ford
thanks: "Code and data are available at: https://github.com/Ford-Robert . Special thanks to Professor Rohan Alexander for motivating this project, providing key support and insight, and demostrating the crucial skills required in modern statistical analysis. Check out his excellent book: https://tellingstorieswithdata.com/"
date: today
date-format: long
abstract: "TODO First sentence. TODO Second sentence. TODO Third sentence. TODO Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

In recent years, Canada's federal spending landscape has undergone significant transformations, particularly with a marked increase in Military spending. The government has announced substantial new military spending packages, [REF]. This paper will use data collected by the Investigative Journalism Foundation (IJF) detailing Canada's Federal Procurement Awards [REF] to analyse government procurements over the last four years.

Federal procurements involve the acquisition of goods and services by government departments, Buyers, from companies, Suppliers. These procurements are essential for a wide range of government functions from; acquiring defense equipment, building infrastructure, procuring healthcare supplies, technological services and engineering consultants to name a few.

@@

I collected and processed data on well over 300,000 contracts awarded since January 2020. The dataset includes variables such as contract amounts, durations, award dates, Buyers, and Suppliers. We derived additional metrics to facilitate our analysis, such as Baseline Days (the number of days since January 1, 2020) and Duration Days (the length of each contract), and standardized supplier names to ensure consistency.

Some significant findings presented in this paper include: military procurement spending has surged in the last two years. Where the Department of National Defence now dominates federal procurement spending, with expenditures exceeding \$56 billion since 2020. This increase in spending is in part thanks to large contracts, including a \$11.2 billion contract to SkyAlyne for fighter pilot training (the largest contract in our dataset) and multi-billion-dollar contracts for fighter jets and naval vessels. While  other government departments have experienced a notable decrease in procurement spending. For instance, the Public Health Agency of Canada's spending, which was substantial in 2020 due to the COVID-19 pandemic, has declined significantly almost nothing in subsequent years.

@@

The paper is structured as follows: @sec-data details the data collection and processing methodologies, including the sources of procurement data and the steps taken to clean and standardize the dataset. Then I present a series of graphs to demonstrate government procurements from various angles to provide a clear picture of spending over the last four years. @sec-model describes the linear regression models used to assess trends in contract amounts over time and contract duration. @sec-results presents the results of the analysis, highlighting key trends in military and non-military procurement spending. @sec-discussion discusses the limitations of the study, including data constraints and methodological considerations, and suggests areas for future research. Finally, @sec-discussion concludes with a summary of the findings and their implications for policymakers, industry stakeholders, and future studies. Additional information about how the data was extracted using web scraping, how the data was cleaned, and how suppliers names were processed can be found in @sec-appendix. Model diagnostics can be found at @sec-diagnostics.

By providing an analysis of federal procurement spending, this paper aims to inform readers on how government consumption of private sector goods and services is changing over time. The insights gained from this study can inform policy decisions, contribute to academic discourse, and enhance transparency regarding government spending practices.

@@

Data was analyzed through the R programming software [@citeR], and `RSelenium` [TODO] with `rvest` [TODO] were used to support a web scraping script to download all data. Additional packages such as `tidyverse` [@tidyverse], `knitr` [@knitr], `actuar` [TODO], `readr` [TODO], `stringr` [TODO], `broom` [TODO] and `dplyr` [@dplyr] were used to help clean, simulate, analyze, and test the data. The presentation of this data using tables and graphs was made possible by r-packages; `ggplot2` [TODO], `lubridate` [TODO], `kableExtra` [TODO], and `zoo` [TODO].

A web scraping script was used to collect all Federal Award Data from the IJF [REF]. The variables this paper focuses are; Contract, Buyer, Supplier, Amount, Award Date, Start Date, and End Date.

@@

The @fig-head1 details the first few rows of the IFJ Database:
```{r fig-head1}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Sample of the Data
#| fig-align: center
#| fig-pos: "H"

@@

Additional Derived Variables, displayed in @fig-head2:

@@

The data was then cleaned. First I began by setting each variable to the correct data type. Then I removed any data point where Amount was negative or zero, because I am not sure what a negative amount means in this context and contracts that are not worth anything are not worth looking at. There were a number of contracts which started and ended on the same day, to simplify later calculation I simply added a day to their End Date, figuring that same day contracts are equivalent to one day contracts. Further data cleaning details are available in @sec-appendix. After cleaning I was left with 320,175 contracts spanning January 2020 to November 2024.

@@

The data is recorded and published through various government agencies, who use different accounting standards and procedures. The three sources IJF used to collect the federal procurement data are; Canada Buys, Buy and Sell, and Proactive Disclosures (which makes up the vast majority of the procurement data). However, because the data used in this paper is pulled from the IJF’s Federal Awards page this data comes exclusively from the Proactive Disclosures. The Treasury Board of Canada Secretariat is responsible for publishing these Proactive Disclosures under the Access to Information Act [REF].

@@

@fig-pie-buyers is a break down of total amount awarded by the top 5 biggest spending Buyers. The Other category consists the of the total awards given by the remaining 88 smallest Government Buyers.  It is clear that National Defence dwarfs the procurement budget, spending over \$56 Billion on procurements since 2020, making up nearly half of all government procurement.

@@

@fig-mil-spending shows how National Defence spending has taken off in recent years, where is 2021 it had not even crested \$5 Billion, but in the last 3 years spending has exploded to \$15 Billion a year. From 2021 to 2022, military procurements increased by more than 300%. Though it is important to note that these amounts reflect the amount awarded in any given year, while many awarded contracts last for years. Especially large contracts which can last decades. Meaning the payouts for these lengthy contracts could be amortized over the length of the contract.

@@

# Create the bar chart
ggplot(data = military_spending_per_year, aes(x = year, y = total_spending)) +
  geom_bar(stat = "identity", fill = def_c) +
  labs(x = "Year", y = "Total Awarded (Billions)") +
  theme_minimal() +
  scale_y_continuous() +
  theme(plot.title = element_text(hjust = 0.5))

@@

Public Services and Procurements Canada (PSPC) is the clearinghouse for the entire federal government. This means they are responsible for managing the procurement process for buyers across the government [REF]. It is important to note that the PSPC sometimes does not publish contracts less than \$40,000. @fig-public-spending shows PSPC spending over time, indicating that general government procurements are taking a hit while military spending balloons.
```{r fig-public-spending}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Public Services Spending Per Year in Billions
#| fig-align: center
#| fig-pos: "H"

@@

# Create the bar chart
ggplot(data = public_spending_per_year, aes(x = year, y = total_spending)) +
  geom_bar(stat = "identity", fill = public_c) +
  labs(x = "Year", y = "Total Awarded (Billions)") +
  theme_minimal() +
  scale_y_continuous() +
  theme(plot.title = element_text(hjust = 0.5))

@@

@fig-top-buyers shows the top five Buyers spending per year since 2020. In the first 2 years military spending held steady and did not dominate the budget by any means. But since then Defence spending exploded not just nominally but compared to the rest of the budget as well. Notably the Public Health Agency of Canada's spending while strong in 2020, decreased to nothing over the following four years. The spending in 2020 is linked to purchases of medical equipment, likely for combating the pandemic. Maybe this over investment has meant that that the Public Health Agency has not needed to invest in equipment in the subsequent years. Though investments across all government agencies, excluding the military, are falling and are lowest in 2024. Note that the 2024 data only goes to November, so more spending may happen in December.

@@

To find out the firms that are benefiting the most through this procurement process we can look to @fig-top-supps. Unsurprisingly, the biggest suppliers are primarily are awarded contracts by the military. Interestingly the top supplier, SkyAlyne, was only awarded one huge \$11.2 Billion contract for fighter pilot training services. This is the largest contract in the dataset. Then considering the largest contracts awarded since 2022 to; the F-35 program (Fighter Planes, \$4.5 Billion), General Atomics (Airplanes, \$1.7 Billion), CAE (Aviation Engineering and Training, \$4.4 Billion), Airbus (Airplanes, \$3.7 Billion), and Textron Aviation (Aircraft parts, \$3.1 Billion ) for a grand total of \$28.6 Billion. These few large contracts devoted to advanced fighter jets, training pilots, and parts are a significant portion of surge in military spending. Remember military procurement spending since 2022 has been about \$15 Billion dollar a year up from \$5 Billion in 2021, as per @fig-mil-spending. That equates to roughly \$30 Billion in extra spending (\$10 Billion extra over the \$5 Billion year for 3 years), \$28.6 Billion of which went to aviation projects.The military also paid Vancouver Shipyards \$3.4 Billion in 2024 for Ships and Boat parts.

@@

Dettwiler is an engineering firm that provides the Canadian Space Agency with engineering and scientific consulting. Surprisingly 86.3% of the Space Agency's procurement budget goes to Dettwiler.

## Contract Distribution

@@

Though most contracts are small contracts (less than \$100,000), the sum of these 246,000 contracts is only worth a measly 7.34 Billion. The SkyAlyne contract itself outweighs all the small contracts.

@@

# Model {#sec-model}

@@

- Tidyverse: For data manipulation and visualization [REF] 
- Dplyr: For data manipulation and transformation [REF] 
- Car: For advanced regression modeling tools and diagnostics [REF] 
- Caret: For classification and regression training, including model tuning and validation [REF]

@@

$$
\text{Amount}_i = \beta_0 + \beta_1 \times \text{Base Line Days}_i + \beta_2 \times \text{Duration Days}_i + \epsilon_i
$$

@@

- Amount: The Amount awarded to a contract
- Baseline Days: The number of days since 1st of January 2020
- Duration Days: The number of days that the contract spans

The point of Baseline Days is to measure when the contracts take place, so when Baseline days is zero then it is one of the earliest contracts, as it gets larger that describes more recent contracts.

@@

# Results {#sec-results}

@@

# Discussion {#sec-discussion}

## What Was Done

@@

Data cleaning was a critical step. Contracts with negative or zero amounts were excluded due to their ambiguous nature. Contracts that started and ended on the same day were adjusted by extending the end date by one day to simplify duration calculations.

@@

## State of Canadian Procurements

@@

@@

# Data Details {#sec-appendix}

## Web Scraping

@@

## Data Cleaning

@@

## Proccessing Supplier

TODO processed_supplier
Talk about what I did thoroughly, see data_cleaning script

@@

# Model Diagnostics {#sec-diagnostics}
